# Odds, Probabilities and Actions

We have come to the point where it will no longer be enough to work with our intuitive notion of probability. If we want to go beyond toy decision problems, like the one's we've been covered, and model decisions that are more like the ones we more frequently face, we'll need to have a sufficiently robust understanding of probability. For example, one of the main things we'll want to be able to do is update our beliefs given new information, which effectively amounts to knowing how to update probabilities.

Probabilities are a way of quantifying beliefs. It may seem impossible to measure something as elusive and subjective as beliefs. But some clever conceptual tools have been developed to do just that.

`r newthought("The key idea behind measuring a person's belief")` about the world is to figure out how willing they are to risk things that they care about. To illustrate this idea we're going to momentarily assume that money is our measure of utility.

## Odds and Fair Betting Rates

Here's a roughly general observation of people's behaviour: the more confident someone is that some event is going to happen, the more willing they are to bet. 

Suppose $S$ is some event that Bob and Ally care about. The event might be a sports team winning a game, that it's going to rain tomorrow, that the stock price of some company will be higher by next year, etc. 

Let's say Bob is more than 50% confident that $S$ will happen. In fact, let's suppose that Bob would accept a deal that would pay him $\$1$ if $S$ happens, but would cost him $\$2$ if it doesn't.   Let's say Ally thinks Bob is wrong to be so confident and agrees to take Bob's bet.  In effect, this means that Ally is willing to put $\$1$ on the table for the chance of winning the $\$2$ that Bob is willing to put on the table. The **stake** is the sum of all the money on the table, in this case $\$1 + \$2 = \$3$. Whoever ends up being right gets to take the stake.^[Careful here. If Ally turns out to win because $S$ does not happen, she wins the stake ($\$3$), but since she contributed $\$1$ to it the amount she *gains* is $\$2$.]

`r newthought("Bob's fair betting rate")` can be expressed by dividing his potential loss by the stake:

$$
  \begin{aligned}
    \mbox{Betting Rate} &= \frac{\mbox{Potential Loss}}{\mbox{Stake}}\\
                        &= \frac{\$2}{\$2 + \$1}\\
                        &= \frac{2}{3}.
  \end{aligned}
$$

Bob's betting rate is $2/3$, which is a reflection of how confident he is that $S$ will happen. And here's the next move: that's Bob's personal probability that $S$ will happen, i.e. $Pr_{Bob}(S)=2/3$. 

`r newthought("Bob's fair odds")` is another way that betting is sometimes talked about. To express Bob's fair better rate in terms of odds, we take the ratio of potential loss to potential win:

$$
  \begin{aligned}
    \mbox{Odds} &= \mbox{Potential Loss : Potential Win}\\
                &=  2:1
  \end{aligned}
$$

The odds that Bob would accept are another reflection of his degree of confidence. In fact, there is a handy way of linking up our notion of expected value with odds and probabilities.  

`r newthought("A fair bet")` is one in which the *expected value* is zero. That is, if we weight the potential win by the probability of winning, and we weight the potential loss by the probability of losing, the odds should "cancel out" or "wash out": $$ (2/3)(\$1) + (1/3)(-\$2) = 0. $$ 

Here's a helpful visual way of understanding the idea of a fair bet.  Notice first an inverse relationship between probabilities and payoffs when it comes to risks (especially in gambling): events with really high payoffs tend to have low probabilities, and likewise, the more probable an event is the lower the payoffs tend to be. If the amount of probability is like the width of a rectangle, and the payoff (or loss) is like the height of a rectangle, then a fair bet will be one in which the area of a rectangle that represents Bob winning will have the same amount of area that represents Bob lossing. 

```{r echo=FALSE, cache=TRUE, fig.margin=FALSE, fig.cap="A bet that pays $\\$1$ if Bob wins and costs $\\$2$ if he loses, is fair when the purple and orange regions have equal area: when the probability of winning is $2/3$."}
library(ggplot2)
library(dplyr)
f <- function(x) case_when(x <= 2/3 ~ 1, x <= 1 ~ 0)
g <- function(x) case_when(x <= 2/3 ~ 0, x <= 1 ~ -2)
ggplot() +
  stat_function(fun = f, geom = "area", n = 1000, fill = "darkviolet") +
  stat_function(fun = g, geom = "area", n = 1000, fill = "darkorange") +
  scale_y_continuous("payoff ($)", breaks = seq(-2, 1, 1), limits = c(-2.5, 1.5)) +
  scale_x_continuous("probability", labels = c("0" = "0", "1/3" = "1/3", "2/3" = "2/3", "1" = "1"), breaks = seq(0, 1, 1/3))
```

To be clear, what Bob considers to be a fair bet might change. For example, suppose Bob comes into some information that significantly decreases his confidence that $S$ will happen. Let's say his confidence goes all the way down to 10% (i.e. 1/10). Since his confidence went down, he should be willing to risk less, that is, he should be willing to stake much less. How much less? As a fair bet, Bob will want to make sure that the expected value will be 0: $$ (1/10)(\$9) + (9/10)(-\$1) = 0. $$ So for Bob to be willing make a bet with Ally give this new information, Ally would need to be willing to put at least $\$9$ in the stake for Bob's $\$1$.

Notice how our visualization using rectangles will change for this new scenario, but the two rectangles will still have the same area.

```{r echo=FALSE, cache=TRUE, fig.margin=FALSE, fig.cap="A bet that pays $\\$9$ if Bob wins and costs $\\$1$ if he loses is fair when the probability of winning is $1/10$."}
f <- function(x) case_when(x <= 1/10 ~ 9, x <= 1 ~ 0)
g <- function(x) case_when(x <= 1/10 ~ 0, x <= 1 ~ -1)
ggplot() +
  stat_function(fun = f, geom = "area", n = 1000, fill = "darkviolet") +
  stat_function(fun = g, geom = "area", n = 1000, fill = "darkorange") +
  scale_y_continuous("payoff ($)", limits = c(-1.5, 10.5)) +
  scale_x_continuous("probability")
```


`r newthought("Here's a General Recipe")` for quantifying a person's probability that a proposition $S$ is true using the idea of fair bets:

1. Find a bet on $S$ that they see as fair. Call the potential winnings $W$ and the potential losses $L$.
2. Because the bet is fair to their eyes, set the expected value of the bet equal to zero:
    $$ [Pr(S) \times W] + [(1-\Pr(S)) \times -L] = 0. $$
3. Now solve for $Pr(S)$:
    $$
      \begin{aligned}
        (Pr(S) \times W)  + ((1-Pr(S)) \times -L) &= 0 \\
        (Pr(S) \times W)  + (-L) + (Pr(S) \times L) &= 0 \\
             (Pr(S) \times W) + (Pr(S) \times L) &= L \\
             Pr(S)\times (W + L) &= L \\
                                       Pr(S) &= \frac{L}{W+L}.
      \end{aligned}
    $$

Notice that we have the formula for the fair betting rate again!  It's helpful to memorize this formula so you don't have to do the derivation each time. But more important than that is knowing that there is a recipe for getting from bets to personal probabilities.


 
