# Conditional Probabilities and Likelihoods

Assuming you have a driver's licence and the roads are relatively clear, the chances of crashing your car are pretty low. But if you're drunk, the chances of crashing are much higher. Probabilities change depending on the conditions.

We already have notation for symbolizing this idea. We use $P(A | B)$ to represent the probability that $A$ is true *given* that $B$ is true. For example, to say the probability of $A$ given $B$ is 30%, we write:
$$ P(A | B) = 0.3 $$
When we condition probabilities in this way, we call them *conditional probabilities*. Conditional probabilities play a central role in the remaining material, so let's spend some time learning how to calculate them.


## Calculating Conditional Probability

```{marginfigure}
Most of this presentation is a light edit of Weisberg's introduction to calculating conditional probability (6.1).
```

```{r echo=FALSE, fig.show='hold', fig.margin=TRUE, fig.cap="Conditional probability in a fair die roll"}
die1 <- readPNG("img/die/die1.png") %>% rasterGrob()
die2 <- readPNG("img/die/die2.png") %>% rasterGrob()
die3 <- readPNG("img/die/die3.png") %>% rasterGrob()
die4 <- readPNG("img/die/die4.png") %>% rasterGrob()
die5 <- readPNG("img/die/die5.png") %>% rasterGrob()
die6 <- readPNG("img/die/die6.png") %>% rasterGrob()
rect1 <- geom_rect(aes(xmin = 0, ymin = 0, xmax = 3, ymax = 1), 
                   size = 1, fill = "darkorange")
rect2 <- geom_rect(aes(xmin = 1.05, ymin = 0.05, xmax = 2.95, ymax = .95), 
                   size = 1, fill = "darkviolet")
p <- ggplot() + 
  theme_void() + coord_fixed() +
  xlim(0, 3) + ylim(0, 2) +
  rect1 +
  annotation_custom(die1, xmin = 0, xmax = 1, ymin = 1, ymax = 2) +
  annotation_custom(die3, xmin = 1, xmax = 2, ymin = 1, ymax = 2) +
  annotation_custom(die5, xmin = 2, xmax = 3, ymin = 1, ymax = 2) +
  annotation_custom(die2, xmin = 0, xmax = 1, ymin = 0, ymax = 1) +
  annotation_custom(die4, xmin = 1, xmax = 2, ymin = 0, ymax = 1) +
  annotation_custom(die6, xmin = 2, xmax = 3, ymin = 0, ymax = 1)
p
p$layers <- append(p$layers, rect2, 3)
p
```

Suppose I roll a fair, six-sided die behind a screen. You can't see the result, but I tell you it's an even number. What's the probability it's also a "high" number: either a $4$, $5$, or $6$?

Maybe you figured the correct answer: $2/3$. But why is that correct? Because, out of the three even numbers ($2$, $4$, and $6$), two of them are high ($4$ and $6$). And since the die is fair, we expect it to land on a high number $2/3$ of the times it lands on an even number.

This hints at a formula for $P(A | B)$.

Conditional Probability

:   
    $$ P(A | B) = \frac{P(A \wedge B)}{P(B)}. $$

In the die-roll example, we considered how many of the $B$ possibilities were also $A$ possibilities. Which means we divided $P(A \wedge B)$ by $P(B)$.

In fact, this formula is our official definition for the concept of conditional probability. When we write the sequence of symbols $P(A | B)$, it's really just shorthand for the fraction $P(A \wedge B) / P(B)$.

```{r condprob, echo=FALSE, fig.margin=TRUE, fig.cap="Conditional probability is the size of the $A \\wedge B$ region compared to the entire $B$ region."}
x <- seq(-.75, .75, 0.01)
upper <- function(x) {
  a <- sqrt(1.5^2 - (x[x < 0] - .75)^2)
  b <- sqrt(1.5^2 - (x[x >= 0] + .75)^2)
  c(a,b)
}
ggplot() + 
  coord_fixed() + theme_void() +
  xlim(-3,3) + ylim(-2,2) +
  geom_circle(aes(x0 = -.75, y0 = 0, r = 1.5), fill = bookorange) +  
  geom_circle(aes(x0 = .75, y0 = 0, r = 1.5), fill = bookblue) +  
  geom_ribbon(aes(x = x, ymin = upper(x), ymax = -upper(x)), 
              fill = bookpurple, colour = "black") +
  geom_text(aes(x = c(-2.25, 2.25), y = c(1, 1), label = c("A", "B")), 
            fontface = "italic", size = 7) +
  theme(panel.border = element_rect(colour = "black", fill = NA, size = 1))
```

In terms of an Euler diagram (Figure \@ref(fig:condprob)), the definition of conditional probability compares the size of the purple $A \wedge B$ region to the size of the whole $B$ region, purple and blue together. If you don't mind getting a little colourful with your algebra:
$$
  P(A | B) = \frac{\color{bookpurple}{\blacksquare}}{\color{bookpurple}{\blacksquare} + \color{bookblue}{\blacksquare}}.
$$
So the definition works because, informally speaking, $P(A \wedge B)/P(B)$ is the proportion of the $B$ outcomes that are also $A$ outcomes.

`r newthought("Dividing")`  by zero is a common pitfall with conditional probability. Notice how the definition of $P(A | B)$ depends on $P(B)$ being larger than zero. If $P(B) = 0$, then the formula
```{marginfigure}
The comedian Steven Wright once quipped that "black holes are where God divided by zero."
```
$$ P(A | B) = \frac{P(A \wedge B)}{P(B)} $$
doesn't even make any sense. There is no number that results from the division on the right hand side.[^alternatesystems]

[^alternatesystems]: There are alternative mathematical systems of probability, where conditional probability is defined differently to avoid this problem. But we'll stick to the standard system. In this system, there's just no such thing as "the probability of $A$ given $B$" when $B$ has zero probability.

In such cases we say that $P(A | B)$ is *undefined*. It's not zero, or some special number. It just isn't a number.

## Application: Monty Hall Problem

Here we'll show how the concept of conditional probability allows us to solve the Monty Hall problem in the introduction. Here's the problem description: 

> On the show there are three doors (A, B, and C), one of which with a prize behind it. You get to pick one of the doors. Let's say you pick A. The host now opens one of the other two doors that you did not pick. But of course, the host doesn't want to give away the game, so the door they open will be empty. After opening one of the two doors (B or C) the host asks, do you want to switch your choice or stick with your current choice of A? 

The intuitive answer, one that many mathematicians and statisticians gave at the time, is that you should be indifferent between switching and staying with your choice of door A. Why? Because, the (incorrect) reasoning goes, there's two doors (A and whichever one the host didn't open) and an even chance between them of where the prize is. Notice that this question is about an *unconditional* probability.

The problem with this reasoning is that it ignores the events that proceeded. The reasoning would be apt if the game show had you picking between just two doors from the very start, and just because the host reveals what's behind the door you didn't pick, they ask you if you want to change your mind. But that's not what's going on the in the Monty Hall problem. The real question is: should you switch your choice from A *given that the host opened a non-prize door after your initial choice*? Notice that this question is about a *conditional* probability.^[What the host does is *not independent* of your initial guess.] 

If the prize is behind door A (the door you initially picked), then the host has a choice between opening up door B or door C. But if the prize isn't behind door A, then the host is constrained. If the prize is behind door B, then the host will open C. If the prize is behind C, then the host will open B. What we're reasoning about here are *paths* of possible events.

The first kind of event is random, it's just about the location of the prize behind one of the three doors. So the probability of the prize being behind door A is 1/3, and similarly for doors B and C. In other words, your initial guess of door A has a 1/3 chance of being right. 

A diagram can help (see Stage 1). Each arrow is a branch and has a probability associated with it.

```{tikz, echo=FALSE, fig.margin = TRUE, fig.cap = "Stage 1 of tree diagram", fig.ext=if (knitr:::is_latex_output()) 'pdf' else 'png'}
\usetikzlibrary{shapes,decorations,arrows,calc,arrows.meta,fit,positioning}
\tikzset{
    -Latex,auto,node distance =1 cm and 1 cm,semithick,
    state/.style ={ellipse, draw, minimum width = 0.7 cm},
    point/.style = {circle, draw, inner sep=0.04cm,fill,node contents={}},
    bidirected/.style={Latex-Latex,dashed},
    el/.style = {inner sep=2pt, align=left, sloped}
}
\begin{tikzpicture}
	\node (x) at (0,0) {};
	\node (a) at (3,3) {A};
	\node (b) at (3,0) {B};
	\node (c) at (3,-3) {C};
	\node (ab) at (6,4) {};
	\node (ac) at (6,2) {};
	\node (bc) at (6,0) {};
	\node (cb) at (6,-3) {};
	\path (x) edge node[] {1/3} (a);
	\path (x) edge node[] {1/3}(b);
	\path (x) edge node[] {1/3}(c); 
\end{tikzpicture}
```
 
The second kind of event is the host's reveal of a non-prize door, which is *not* random *if* your choice of door A is *incorrect*, and is random if your choice of A is correct. Let's build on the diagram above, showing what the host's options are (see Stage 2). When the prize is behind door A (your guess) the host has two options, so we have two branches, each with a probability of 1/2. When the prize is behind door B, there's only one thing the host can do (so that branch has a probability of 1, which we don't bother labeling). Same thing when the prize is behind door C.

```{tikz, echo=FALSE, fig.margin = TRUE, fig.cap = "Stage 2 of tree diagram", fig.ext=if (knitr:::is_latex_output()) 'pdf' else 'png'}
\usetikzlibrary{shapes,decorations,arrows,calc,arrows.meta,fit,positioning}
\tikzset{
    -Latex,auto,node distance =1 cm and 1 cm,semithick,
    state/.style ={ellipse, draw, minimum width = 0.7 cm},
    point/.style = {circle, draw, inner sep=0.04cm,fill,node contents={}},
    bidirected/.style={Latex-Latex,dashed},
    el/.style = {inner sep=2pt, align=left, sloped}
}
\begin{tikzpicture}
	\node (x) at (0,0) {};
	\node (a) at (3,3) {A};
	\node (b) at (3,0) {B};
	\node (c) at (3,-3) {C};
	\node (ab) at (6,4) {Open B };
	\node (ac) at (6,2) {Open C};
	\node (bc) at (6,0) {Open C};
	\node (cb) at (6,-3) {Open B};
	\path (x) edge node[] {1/3} (a);
	\path (x) edge node[] {1/3} (b);
	\path (x) edge node[] {1/3} (c);
	\path (a) edge node[] {1/2} (ab);
	\path (a) edge node[] {1/2} (ac);
	\path (b) edge (bc);
	\path (c) edge (cb);
\end{tikzpicture}
```

Now imagine you could play this game over and over again (thousands if you like!), always making your first pick door A.^[You can try it for yourself [HERE.](http://www.rossmanchance.com/applets/2021/montyhall/Monty.html)] One third of the time the prize will be behind door A, and of those times, half of them the host opens door B and the other half door C. So that means that the top path where the prize is behind door A and the host opens door B will happen 1/6th of the time. Similarly for the path where the prize is behind door A and the host opens C. For the other two possibilities where the prize is behind door B or door C, the host only has one option, so each of those paths will happen 1/3rd of the time. Here's our completed diagram with the probabilities at the end of the paths.

```{tikz, echo=FALSE, fig.cap = "Stage 3 of tree diagram", fig.ext=if (knitr:::is_latex_output()) 'pdf' else 'png'}
\usetikzlibrary{shapes,decorations,arrows,calc,arrows.meta,fit,positioning}
\tikzset{
    -Latex,auto,node distance =1 cm and 1 cm,semithick,
    state/.style ={ellipse, draw, minimum width = 0.7 cm},
    point/.style = {circle, draw, inner sep=0.04cm,fill,node contents={}},
    bidirected/.style={Latex-Latex,dashed},
    el/.style = {inner sep=2pt, align=left, sloped}
}
\begin{tikzpicture}
	\node (x) at (0,0) {};
	\node (a) at (3,3) {A};
	\node (b) at (3,0) {B};
	\node (c) at (3,-3) {C};
	\node (ab) at (6,4) {Open B };
	\node (ac) at (6,2) {Open C};
	\node (bc) at (6,0) {Open C};
	\node (cb) at (6,-3) {Open B};
	\node (abp) at (8,4) {p=1/6};
	\node (acp) at (8,2) {p=1/6};
	\node (bcp) at (8,0) {p=1/3};
	\node (cbp) at (8,-3) {p=1/3};		
	\path (x) edge node[] {1/3} (a);
	\path (x) edge node[] {1/3} (b);
	\path (x) edge node[] {1/3} (c);
	\path (a) edge node[] {1/2} (ab);
	\path (a) edge node[] {1/2} (ac);
	\path (b) edge (bc);
	\path (c) edge (cb);
\end{tikzpicture}
```

Of course you don't actually know which path you are on, but here's what you do know: When the host opens door C, you are more likely to be on the middle path (1/3) than you are on the path above it (1/6). On the middle path the prize is behind door B, while in the path above the prize is behind door A. Since the middle path is the more likely one you're on, you should switch your choice from A to B! The same reasoning applies if the host had opened door B instead - then you should switch your choice to C.

Notice that the probabilities at the end of the paths have to taken into account all the branches that lead to that end point. So the probability of being on the branch that goes from A to Open B is *conditional* on the previous branch that goes to A.


Here's another example.^[Thanks to Weisberg's *Odds and Ends*.]

Suppose there are two urns of coloured marbles.

- Urn X contains 3 black marbles, 1 white.
- Urn Y contains 1 black marble, 3 white.

We flip a fair coin to decide which urn to draw from, heads for Urn X and tails for Urn Y. Then we draw one marble at random. The following tree diagram shows us the possible sequences of events.


```{tikz, echo=FALSE, fig.margin = TRUE, fig.cap = "Coin Flip and Urn Draws", fig.ext=if (knitr:::is_latex_output()) 'pdf' else 'png'}
\usetikzlibrary{shapes,decorations,arrows,calc,arrows.meta,fit,positioning}
\tikzset{
    -Latex,auto,node distance =1 cm and 1 cm,semithick,
    state/.style ={ellipse, draw, minimum width = 0.7 cm},
    point/.style = {circle, draw, inner sep=0.04cm,fill,node contents={}},
    bidirected/.style={Latex-Latex,dashed},
    el/.style = {inner sep=2pt, align=left, sloped}
}
\begin{tikzpicture}
	\node (x) at (0,0){};
	\node (h) at (3,2){H};
	\node (t) at (3,-2){T};
	\node (hb) at (6,3) {B};
	\node (hw) at (6,1) {W};
	\node (tb) at (6,-1) {B};
	\node (tw) at (6,-3) {W};
	\node (hbp) at (7,3) {3/8};
	\node (hwp) at (7,1) {1/8};
	\node (tbp) at (7,-1) {1/8};
	\node (twp) at (7,-3) {3/8};
	\path (x) edge node[] {1/2} (h);
	\path (x) edge node[] {1/2} (t);
	\path (h) edge node[] {3/4} (hb);
	\path (h) edge node[] {1/4} (hw);
	\path (t) edge node[] {1/4} (tb);
	\path (t) edge node[] {3/4} (tw);
\end{tikzpicture}
```

The probability of drawing a black marble on the top path is $3/4$ because we are assuming the coin landed heads, and thus we're drawing from Urn X. If the coin lands tails instead, and we draw from Urn Y, then the chance of a black marble is instead $1/4$. So these quantities are conditional probabilities:
$$
  \begin{aligned}
    Pr(B | H) &= 3/4,\\
    Pr(B | T) &= 1/4.
  \end{aligned}
$$
Notice, though, the first branch in a tree diagram is different. In the $H$-vs.-$T$ branch, the probabilities are *un*conditional, since there are no previous branches for them to be conditional on.

## Independence

`r newthought("We can state independence")` in a rigorous way now. Informally stated, we said that A and B are independent when the probability of B (or the truth of B) does not affect the probability of A. The concept of conditional probability allows us to say this explicitly using probability theory:

Independence

:    $A$ is independent of $B$ if $Pr(A | B) = Pr(A)$ and $Pr(A) > 0$.


We will see more of this concept of independence, as it plays an important role in calculating probabilities. But first, there is an important kind of conditional probability called a *likelihood* that is worth taking some time to explain. 


## Likelihoods

`r newthought("Order matters")` when it comes to conditional probabilities. For example, given that someone is a university student (S), the probability that they are below 40 years old (F) is relatively high. However, given that someone is below 40 years old, the probability that they are a university student is low. In symbols, $Pr(S|F)\neq Pr(F|S)$.^[There are some cases where the two are the same, but that's rare.]

`r newthought("Here's another example.")`^[Borrowed from Weisberg's *Odds and Ends*, chapter 6.] Suppose a university has 10,000 students. Each is studying under one of four broad headings: Humanities, Social Sciences, STEM, or Professional. Under each of these categories, the number of students with an average grade of A, B, C, or D is listed in the following table. What is the probability a randomly selected student will have an A average, given that they are studying either Humanities or Social Sciences?

```{r studentgrades, echo=FALSE, echo=FALSE}
df <- data.frame(
    `Humanities`      = c(200, 500, 250, 50),
    `Social Sciences` = c(600, 800, 400, 200),
    `STEM`            = c(400, 1600, 1500, 500),
    `Professional`    = c(900, 900, 750, 450)
)
colnames(df) <- c("Humanities", "Social Sciences", "STEM", "Professional")
rownames(df) <- c("A", "B", "C", "D")
knitr::kable(df, align = "c")
```

$$
  \begin{aligned}
    Pr(A ~ | ~(H \vee S)~) &= \frac{Pr(A \wedge (H \vee S))}{Pr(H \vee S)}\\
                           &= \frac{800/10,000}{3,000/10,000}\\
                           &= 4/15.
  \end{aligned}
$$
What about the reverse probability, that a student is studying either Humanities or Social Sciences given that they have an A average?
$$
  \begin{aligned}
    Pr((H \vee S) ~ | ~ A) &= \frac{Pr((H \vee S) \wedge A)}{Pr(A)}\\
                           &= \frac{800/10,000}{2,100/10,000}\\
                           &= 8/21.
  \end{aligned}
$$
Notice how we get a different number now.

`r newthought("When we reason about hypotheses")` the order is particularly important to get right, so much so that there is a particular term that scientists will use to indicate the direction of the conditional probability. This is called a *likelihood*. To understand it, we need to make an important distinction between a hypothesis (or theory) and evidence (or data or observation).  

`r newthought("For example,")` suppose a company BestShoes claims that 96% of their shoes outlast their competitor's average shoe mileage of 300 miles. Furthermore, lets say that a running magazine decides to test BestShoes' claim by sending out 2,400 shoes from BestShoes to the magazine's subscribers. The magazine finds that 133 of the shoes did not make it past 300 miles before falling apart. The *hypothesis* is the claim that each shoe has a 0.96 probability of making it past 300 miles. The *evidence* is the claim that 2,277 of 2,400 shoes made it past 300 miles.^[Later we'll see how to go about quantifying the degree of support (or lack there of) that the evidence gives to the hypothesis.] 

It will be convenient for us to use the letter *H* when talking about some hypothesis and *E* for some statement of evidence.

In general, a *hypothesis* is a statement about what is (or will be) the case. Our beliefs about the world are similar to scientific hypotheses (though perhaps not as technically stated) in the following sense: 

  - i) there is the content of our belief, which is the statement (e.g. that BestShoes outlast their competitors), and then there is 
  - ii) the degree of confidence that the statement is true, which we previously learned to (indirectly) measure with the concept of fair bets. We have been representing this degree of confidence as a (personal) probability that a proposition is true, e.g. $Pr(H)=0.75$ means a person is willing to take 3:1 odds in favor that $H$ is true.

The idea of *confidence* as a degree of belief is not to be confused with *reasons* for that level of confidence.   *Evidence* is a kind of reason that can be used to *support* a hypothesis. Evidence gives us reasons for believing that a hypothesis is true: more evidence should make us more confident. 

The relationship between hypothesis and evidence can be tricky, in part because there are two different directions that are not equivalent. Recall that most of the time $P(A|B)\neq P(B|A)$. That same lesson holds for hypothesis and evidence: $P(H|E)\neq P(E|H)$.^[In words: the probability that a hypothesis is true *given* that the evidence is true *is not equal to* the probability that the evidence is true given that the hypothesis is true.]

Consider the following two conditional statements:

1. If we have the body of evidence $E$, how probable is it that hypothesis $H$ is true?
2. If hypothesis $H$ is true, then how likely is it that we would see the body of evidence $E$?

It is a bit of curious fact that a great deal of statistics focuses on developing tools for understanding claim (2) when what we typically want to know is claim (1). For example, in most statistics classes that focus on hypothesis testing, the running intuition goes something like this:

> If the hypothesis we're testing is true, then how unusual would this result (the evidence) be?

The core idea behind all the statistical machinery (specifically what's called *frequentist* statistics) is that if the evidence would be unusual below some threshold level (called $\alpha$ - "alpha") *under the hypothetical assumption that the hypothesis were true*, then that supports or gives us reason to think that the hypothesis is true. If that sounds confusing, it's because it is. What we're trying to say is something like, "if the hypothesis were true, then this evidence we're seeing would be *too much of a coincidence*". If that doesn't help, you're not alone. Even scientists and statisticians have been found to misunderstand the idea.^[See https://jonathanweisberg.org/vip/significance-testing.html#significance-testing for an excellent introduction with examples of hypothesis testing.]

The point we need to make for our purposes is this. When scientists talk about likelihoods, they don't mean just any kind of probability. What they mean is as follows.

`r newthought("A likelihood")` is a conditional probability of seeing some evidence given the assumption that a hypothesis is true, i.e. $P(E|H)$.


## Application: The Taxi Cab Problem

If there's anything to take away from this chapter about how conditional probabilities are used in (scientific) reasoning, it's to make sure you remember that  $P(E|H)\neq P(H|E)$.

`r newthought("To illustrate")` this point, consider the following famous taxi cab problem.
```{marginfigure}
The experiment was first published in 1971. It was performed by [Daniel Kahneman](https://en.wikipedia.org/wiki/Daniel_Kahneman) and [Amos Tversky](https://en.wikipedia.org/wiki/Amos_Tversky). Their work on human reasoning reshaped the field of psychology, and eventually won a Nobel prize in 2002.
```

A cab was involved in a hit and run accident at night. Two cab companies, the Green and the Blue, operate in the city. You are given the following data:

1.  $85\%$ of the cabs in the city are Green and $15\%$ are Blue.
2.  A witness identified the cab as Blue. The court tested the reliability of the witness under the same circumstances that existed on the night of the accident and concluded that the witness correctly identified each one of the two colors $80\%$ of the time and failed $20\%$ of the time.

What is the probability that the cab involved in the accident was blue rather green?


Most people answer $80\%$, because the witness is $80\%$ reliable. But the right answer is $12/29$, or about $41\%$.

How could the probability be so low when the witness is $80\%$ reliable? The short answer is: because blue cabs are rare. So most of the time, when the witness says a cab is blue, it's one of the $20\%$ of green cabs they mistakenly identify as blue.

A diagram can help.

```{r taxigrid, echo=FALSE, fig.margin=TRUE, fig.cap="The taxicab problem. There are $15$ blue cabs, $85$ green. The dashed region indicates those cabs the witness identifies as \"blue.\" It includes $80\\%$ of the blue cabs ($12$), and only $20\\%$ of the green ones ($17$). Yet it includes more green cabs than blue."}
taxi_blue <- readPNG("img/taxi_blue.png") %>% rasterGrob()
taxi_green <- readPNG("img/taxi_green.png") %>% rasterGrob()
place_image <- function(row, img) {
  annotation_custom(img, xmin = row[1], xmax = row[1] + 1, 
                    ymin = row[2] + .1, ymax = row[2] + .9)
}
green_grid <- bind_rows(
  expand.grid(x = 0:9, y = 0:6),
  expand.grid(x = 5:9, y = 7:9)
)
blue_grid <- expand.grid(x = 0:4, y = 7:9)
segments <- bind_rows(
  list(x = 0, y = 2, xend = 0, yend = 10),
  list(x = 0, y = 10, xend = 4, yend = 10),
  list(x = 4, y = 10, xend = 4, yend = 3),
  list(x = 4, y = 3, xend = 1, yend = 3),
  list(x = 1, y = 3, xend = 1, yend = 2),
  list(x = 1, y = 3, xend = 1, yend = 2),
  list(x = 1, y = 2, xend = 0, yend = 2)
)
ggplot() + 
  theme_void() + xlim(0, 10) + ylim(0, 10) +
  apply(green_grid, 1, place_image, taxi_green) +
  apply(blue_grid, 1, place_image, taxi_blue) +
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend),
               data = segments, colour = "blue", size = 1, linetype = "longdash")
```


Imagine there are just $100$ cabs in town, $85$ green and $15$ blue. The dashed blue line represents the cabs the witness identifies as 
"blue," both right or wrong. Because the witness is $80\%$ accurate, that line encompasses $80\%$ of the blue cabs, which is $12$ cabs. But it also encompasses $20\%$ of the green cabs, which is $17$. That's a total of $29$ cabs identified as "blue," only $12$ of which actually are blue.

So out of the $29$ cabs the witness calls "blue," only $12$ really are blue. The probability a cab really is blue given the witness says so is only $12/29$, about $41\%$.

Another way to think about the problem is that there are *two* pieces of information relevant to whether the cab is blue. The witness says the cab is blue, but also, most cabs are not blue. So there's evidence for the cab being blue, but also strong evidence against it. The diagram shows us how to balance these two, competing pieces of evidence and come to the correct answer.

What trips people up so much in the taxicab problem? Remember how order matters with conditional probability. In this problem, we're asked to find $Pr(B | W)$, the probability the cab is blue given that the witness says it is. That's not the same as $Pr(W | B)$, the probability the witness will say the cab is blue if it really is. The problem tells us $Pr(W | B) = 8/10$, but it doesn't tell us a number for $Pr(B | W)$. We have to figure that out.


`r newthought("It's not that the two conditional probabilities")` aren't related. The taxi cab problem highlights why we should keep the two distinct, but we'd be wrong to think that they aren't related at all. In fact a very famous theorem called *Bayes Theorem* shows us a fundamental connection between them. We show that next.


## Exercises

1. Answer each of the following:

    ```{marginfigure, echo=TRUE}
    Nearly all of these exercises come directly from or are versions of those in Chapter 6 of Weisberg's *Odds and Ends*.
    ```

    a.  On a fair die with six sides, what is the probability of rolling a low number (1, 2, or 3) given that you roll an even number.
    a.  On a fair die with eight sides, what is the probability of rolling an even number given that you roll a high number (5, 6, 7, or 8)?

1. Suppose $Pr(B) = 4/10$, $Pr(A) = 7/10$, and $Pr(B \wedge A) = 2/10$.
    What are each of the following probabilities?

    a.  $Pr(A | B)$
    a.  $Pr(B | A)$

1. Five percent of tablets made by the company Ixian have factory defects. Ten percent of the tablets made by their competitor company Guild do. A computer store buys $40\%$ of its tablets from Ixian, and $60\%$ from Guild.

    ```{marginfigure, echo=TRUE}
    This exercise and the next one are based on very similar exercises from Ian Hacking's wonderful book, *An Introduction to Probability and Inductive Logic*.
    ```

    Draw a probability tree to answer the following questions.

    a. What is the probability a randomly selected tablet in the store is made by Ixian and has a factory defect?
    #. What is the probability a randomly selected tablet in the store has a factory defect?
    #. What is the probability a tablet from this store is made by Ixian, given that it has a factory defect?

1. In the city of Elizabeth, the neighbourhood of Southside has lots of chemical plants. $2\%$ of Elizabeth's children live in Southside, and $14\%$ of those children have been exposed to toxic levels of lead. Elsewhere in the city, only $1\%$ of the children have toxic levels of exposure.

    Draw a probability tree to answer the following questions.

    a. What is the probability that a randomly chosen child from Elizabeth  lives in Southside and has toxic levels of lead exposure?
    b. What is the probability that a randomly chosen child from Elizabeth has toxic levels of lead exposure?
    c. What is the probability that a randomly chosen child from Elizabeth who has toxic levels of lead exposure lives in Southside?

1. Imagine 100 prisoners are sentenced to death. 70 of them are housed in cell block A, the other 30 are in cell block B. Of the prisoners in cell block A, 9 are innocent. Only 1 prisoner in cell block B is innocent.

    The law requires that one prisoner be pardoned. The lucky prisoner will be selected by flipping a fair coin to choose either cell block A or B. Then a fair lottery will be used to select a random prisoner from the chosen cell block.

    What is the probability the pardoned prisoner comes from cell block A if she is innocent? Answer each of the following to find out.

    $I$ = The pardoned prisoner is innocent.\
    $A$ = The pardoned prisoner comes from cell block A.

    a.  What is $Pr(I | A)$?
    b.  What is $Pr(A \wedge I)$?
    c.  What is $Pr(I | B)$?
    d.  What is $Pr(B \wedge I)$?
    e.  What is $Pr(I)$?
    f.  What is $Pr(A | I)$?
    g.  Draw a probability tree to visualize and verify your calculations.
    
1. Suppose $A$, $B$, and $C$ are independent, and they each have the same probability: $1/3$. What is $Pr(A \wedge B | C)$?
    
    
1. If $A$ and $B$ are mutually exclusive, what is $Pr(A | B)$? Justify your answer using the definition of conditional probability.    
    
1.  Which of the following situations is impossible? Justify your answer.

    a.  $Pr(A) = 1/2, Pr(A | B) = 1/2, Pr(B | A) = 1/2$.
    b.  $Pr(A) = 1/2, Pr(A | B) = 1, Pr(A | \neg B) = 1$.

1. Is the following statement true or false: if $A$ and $B$ are mutually exclusive, then $Pr(A \vee B | C) = Pr(A | C) + Pr(B | C)$. Justify your answer.

1. Justify the second part of the "independence means multiply" mantra: if $Pr(A) > 0$, $Pr(B) > 0$, and $Pr(A \wedge B) = Pr(A) Pr(B)$, then $A$ is independent of $B$.

    ```{marginfigure, echo=TRUE}
    This and the remaining exercises are more challenging and ask you to think more deeply about conditional probability and independence.
    ```

    Hint: start by supposing $Pr(A) > 0$, $Pr(B) > 0$, and $Pr(A \wedge B) = Pr(A)Pr(B)$. Then apply some algebra and the definition of conditional probability.

1. Justify the claim that independence is symmetric: if $A$ is independent of $B$, then $B$ is independent of $A$.

    Hint: start by supposing that $A$ is independent of $B$. Then write out $Pr(A | B)$ and apply the definition of conditional probability.

1. Suppose $A$, $B$, and $C$ are independent. Is it possible that $Pr(A \wedge B \wedge C) = 0$? If yes, give an example where this happens. If no, prove that it cannot happen.

1. Suppose we have $4$ apples and $10$ buckets. We place each apple in a random bucket; the placement of each apple is independent of the others. Let $B_{ij}$ be the proposition that apples $i$ and $j$ were placed in the same bucket.

    a. Is $B_{12}$ independent of $B_{34}$?
    a. Is $B_{12}$ independent of $B_{23}$?
    a. Is every pair of $B_{ij}$ propositions independent?
    a. Is every trio of $B_{ij}$ propositions independent?

1.  Suppose we have a coin whose bias we want to learn, so we're going to flip it $3$ times. We start out by assigning the same probability to each possible sequence of heads and tails. For example, the sequences $HTH$ and $TTT$ are equally likely, as are all other sequences. 

    a. Before we do our $3$ flips, what is the probability of $HTH$?
    a. What is the probability of heads on the third flip, given that the first two flips land heads?

1. Prove that if $A$ logically entails $B$, then $Pr(B | A) = 1$.

1. Suppose the following three conditions hold:

    i.   $Pr(A) = Pr(\neg A)$,
    ii.  $Pr(B | A) = Pr(B | \neg A)$,
    iii. $Pr(B) > 0$.
    
    Must the following be true then?
    
    iv.  $Pr(A | B) = Pr(A | \neg B) = 1/2$?
    
    If yes, prove that (iv) must hold. If no, give a counterexample: draw an Euler diagram where conditions (i)--(iii) hold, but not (iv).

1. Prove that the equation $Pr(A | B) Pr(B) = Pr(B | A) Pr(A)$ always holds. (Assume both conditional probabilities are well-defined.)

1. Prove that the following equation always holds, assuming the conditional probabilities are well-defined:
    $$ \frac{Pr(A | B)}{Pr(B | A)} = \frac{Pr(A)}{Pr(B)}. $$

1. Does the equation $Pr(A | B) = Pr(\neg B | \neg A)$ always hold, assuming both conditional probabilities are both well-defined? If yes, prove that it does. If no, draw an eikosogram where it fails to hold.

1. Suppose an urn contains 30 black marbles and 70 white. We randomly draw 5 marbles with replacement. Let $A$ be the proposition that $3$ of the first $4$ draws are black, and let $B$ be the proposition that the $5$^th^ draw is black. Calculate the following quantity:
    $$ \frac{Pr(A \,\vert\, B)}{Pr(B \,\vert\, A)} \frac{Pr(B)}{Pr(A)}. $$


<!-- 
Use two lotteries (or slot machines) as examples.

-->

<!-- Look at Lindley's presentation of exchangeability? (Maybe save this until the learning chapter? Or maybe do one iteration of it here, then return to it again.) -->

